Using the internal language of toposes in algebraic geometry

### 2016-05-24 slides-cambridge2016.tex
Updated note slides

\documentclass[12pt,utf8,notheorems,compress,t,draft]{beamer}
\usepackage{etex}

@@

\IfSubStr{\jobname}{\detokenize{nonotes}}{
  \setbeameroption{hide notes}
}{
  \setbeameroption{show notes}
}
\setbeamertemplate{note page}[plain]
\setbeameroption{hide notes}  % XXX

@@

@@

{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/cambridge-university-botanic-garden}}
\begin{frame}[plain]
  \vspace{\textheight}\vspace{-1em}
  \centering
  \emph{Cambridge University Botanic Garden} \\
  \par
\end{frame}}

@@

\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}

@@

  \begin{changemargin}{2.5em}{2.5em}
    % XXX: Update
    We describe how the internal language of certain
    toposes, the associated small and big Zariski toposes of a scheme, can be
    used to give simpler definitions and more conceptual proofs of the basic
    notions and observations in algebraic geometry.
    % This is useful for studying schemes from a local or relative point of view.

@@

    A particularly interesting task is to internalize the
    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
    view, this construction should simply reduce to an intuitionistically sensible
    variant of the ordinary construction of the spectrum of a ring, but it turns
    out that this expectation is too naive and that a refined approach is
    necessary.
  \end{changemargin}
}

@@

  A sheaf~$\O_X$ of rings is \emph{local} if and only if all the
  stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
  sections~$\O_X(U)$ are local rings? This has a geometric meaning, but can
  also be motivated from a logical point of view: A sheaf of rings is local if
  and only if, from the point of view of the internal language of~$\Sh(X)$, it
  is a local ring.

@@

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\E$ \\
      \midrule
      object of~$\E$ & set/type \\
      morphism in~$\E$ & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      group object & group \\
      \bottomrule
    \end{tabular}
  \end{center}

@@

      \item \begin{justify}For expositions of the internal language,
      see the books~\cite{johnstone:elephant}~(chapters~D1 and~D4)
      and~\cite{moerdijk:maclane:sheaves} (chapter~VI) or the lecture
      notes~\cite{streicher:ctcl} (chapter~13).\end{justify}
    \end{itemize}
  }
}

\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
  Let~$X$ be a topological space. Then we recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  \pause
  \small
  \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}l@{\ }c@{\ }l@{}}
    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
    U \models \varphi \wedge \psi &\Ll&
      \text{$U \models \varphi$ and $U \models \psi$} \\
    U \models \varphi \vee \psi &\Ll&
      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
    U \models \varphi \Rightarrow \psi &\Ll&
      \text{for all open~$V \subseteq U$: } 
    \text{$V \models \varphi$ implies $V \models \psi$} \\
    U \models \forall f \? \F\_ \varphi(f) &\Ll&
      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
      \varphi(f)$} \\
    U \models \exists f \? \F\_ \varphi(f) &\Ll&
      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
    $U_i \models \varphi(f_i)$}
  \end{array} \]

  \note{
    \begin{itemize}
      \item \begin{justify}The rules are called \emph{Kripke-Joyal semantics} and can be
      formulated over any topos (not just sheaf topoi). They are not all
      arbitrary: Rather, they are very finely concerted to make the crucial
      properties about the internal language (see next slide) true.\end{justify}

@@

      \item Negation is defined as
      \[ \neg\varphi :\equiv (\varphi \Rightarrow \bot). \]

      \item \begin{justify}The alternate definition ``$U \models \varphi \vee \psi :\Leftrightarrow
      \text{$U \models \varphi$ or $U \models \psi$}$'' would not be local (cf.
      next slide).\end{justify}
    \end{itemize}
  }
}

@@

\note{
  \begin{itemize}
    \item One can simplify the rules for often-occuring special cases:
    \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
      U \models \forall s\?\F\_ \forall t\?\G\_ \varphi(s,t)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$,} \\
      && \text{sections~$s \in \F(V)$, $t \in \G(V)$:} \\
      &&\qquad\qquad V \models \varphi(s,t) \\\\
      U \models \forall s\?\F\_ \varphi(s) \Rightarrow \psi(s)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$, sections~$s \in \F(V)$:} \\
      &&\qquad\qquad \text{$V \models \varphi(s)$ implies $V \models \psi(s)$} \\\\
      U \models \exists!s\?\F\_ \varphi(s)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$,} \\
      &&
      \text{there is exactly one section~$s \in \F(V)$ with:} \\
      &&\qquad\qquad V \models \varphi(s)
    \end{array} \]
  \end{itemize}
}

\note{
  \begin{itemize}
    \item \begin{justify}One can extend the language to allow for \emph{unbounded}
    quantification ($\forall A$ vs.~$\forall a \in A$), see Shulman's
    stack semantics~\cite{shulman:stack}. This is needed to formulate
    universal properties internal to~$\Sh(X)$, for instance.\end{justify}

    \item \begin{justify}One can further extend the language to be able to talk
    about locally internal categories over~$\Sh(X)$ (in the sense of
    Penon~\cite{penon:locint}, see for instance the appendix
    of~\cite{johnstone:topos}): Then one can do category theory internal
    to~$\Sh(X)$ using the internal language, see~\cite{blechschmidt:msc}.

    This specific approach is, as far as I am aware, original work. But of
    course, internal category theory has been done for a long time, see for
    instance the textbook~\cite{johnstone:elephant}; also
    cf.~\cite{chapman:rowbottom:relcat}.\end{justify}
  \end{itemize}
}

@@

\note{%
  \vspace{-0.5em}%
  \begin{center}\large\bf Why is constructive mathematics
  interesting?\end{center}
  \vspace{-1em}
  \begin{itemize}
    \item The internal logic of most topoi is constructive.

@@

\note{%
  \vspace{-0.5em}%
  \begin{center}\large\bf Three rumours about constructive mathematics\end{center}
  \vspace{-1em}

  \begin{enumerate}
    \item \begin{justify}There is a false rumour about constructive mathematics, namely that
    the term \emph{contradiction} is generally forbidden. This is not the
    case, one has to distinguish between
    \begin{itemize}
      \item a true proof by contradiction: ``Assume~$\varphi$ were false.
      Then \ldots, contradiction. So~$\varphi$ is in fact true.''
    \end{itemize}
    which constructively is only a proof of the weaker
    statement~$\neg\neg\varphi$, and
    \begin{itemize}
      \item a proof of a negated formula: ``Assume~$\psi$ were true. Then
      \ldots, contradiction. So~$\neg\psi$ holds.''
    \end{itemize}
    which is a perfectly fine proof of~$\neg\psi$ in constructive mathematics.\end{justify}

@@

\note{
  \begin{enumerate}
    \item[3.] \begin{justify}There is one last false rumour about constructive mathematics: Namely
    that most of mathematics breaks down in a constructive setting. This is
    only true if interpreted naively: Often, already very small changes to the
    definitions and statements (which are classically simply equivalent
    reformulations) suffice to make them constructively acceptable.

    In other cases, adding an additional hypotheses, which is classically
    always satisfied, is necessary (and interesting). Here is an example: In
    constructive mathematics, one can not show that any inhabited subset of the
    natural numbers possesses a minimal element. [One can also not show the
    negation -- recall the previous false rumour.] But one can show (quite
    easily, by induction) that any inhabited and \emph{detachable} subset
    of the natural numbers possesses a minimal element. A subset~$U \subseteq
    \NN$ is detachable iff for any number~$n \in \NN$, it holds that~$n
    \in U$ or~$n \not\in U$.

@@

  \begin{justify}See~\cite{stanford:constr,dalen:int,troelstra:dalen:constr}
  for references about constructive mathematics. The blog~\cite{bauer:blog} is
  also very informative.\end{justify}
}

@@

\note{\begin{itemize}\justifying
  \item Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
  For instance, there is no global membership relation~$\in$. Rather,
  for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
  \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
  object of truth values of~$\E$ (can be understood as the power object of
  a terminal object).

  \item Compare with the embedding theorem for abelian categories:
  There, an explicit embedding into a category of modules is constructed.
  Here, we only change perspective and talk about the same objects and
  morphisms.

  \item There exists a weaker variant of the internal language which works
  in abelian categories. By using it, one can even pretend that the objects
  are abelian groups (instead of modules), and when constructing morphisms
  by appealing to the axiom of unique choice (which is a theorem), one
  doesn't even have to check linearity. The proof that this approach
  works uses only categorical logic.
\end{itemize}}

@@

\begin{frame}\frametitle{Building a dictionary}
  \sloganwithoutborder{\hil{Understand notions of algebraic geometry
  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \scalebox{0.93}{\begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set/type \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      finite locally free sheaf & finite free module \\
      % coherent sheaf & coherent module \\
      tensor product of sheaves & tensor product of modules \\
      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
      sheaf of Kähler differentials & module of Kähler differentials \\
      sheaf of rational functions & total quotient ring of~$\O_X$ \\
      \bottomrule
    \end{tabular}}
  \end{center}

@@

@@

  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
  \begin{itemize}
  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic filter}.
  \end{itemize}
  Note: $M$ and $\underline{M}$ share all first-order properties.
  \pause
  \vfill

@@

@@

\begin{frame}\frametitle{The curious role of affine open subsets}
  \hil{Question:} Why do the following identities hold, for quasicoherent
  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
  \begin{align*}
    \E/\F(U) &= \E(U) / \F(U) \\
    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
  \end{align*}
  \pause
  \hil{A calculation:}
  \begin{align*}
    M^\sim \otimes_{\O_X} N^\sim &=
    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
    &= (\ull{M \otimes_A N})[\F^{-1}] =
    (M \otimes_A N)^\sim.
  \end{align*}
  \pause
  \hil{Answer:} Because localization commutes with quotients, tensor products,
  torsion submodules (sometimes), \ldots
\end{frame}

@@

  Then~$\E$ is quasicoherent
  if and only if, internally to~$\Sh(X)$,
  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Diamond_f$-sheaf for any~$f : \O_X$, \\[0.3em]
  \qquad\qquad where~$\Diamond_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
  \end{quote}
  \pause

@@

\note{\justifying
  The sheaf condition and the sheafification functor can be described purely
  internally. An object~$M$ is \emph{separated} with respect to~$\Diamond$ if
  and only if, from the internal point of view,
  \[ \forall x,y : M\_\ \Diamond(x = y) \Rightarrow x = y. \]
  It is a \emph{sheaf} with respect to~$\Diamond$, if furthermore
  \[ \forall K \subseteq M\_\
    \Diamond(\exists x : M\_\ K = \{ x \}) \Longrightarrow
    \exists x : M\_\ \Diamond(x \in K). \]

@@

\newcommand{\idiamond}{{\usebeamercolor[fg]{item}{\boldsymbol{\Diamond}}}}
\newcommand{\gdiamond}[1]{\textcolor{gray}{\boldsymbol{\Diamond}(}#1\textcolor{gray}{)}}

>>>>

\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
\usepackage{etex}

@@

\IfSubStr{\jobname}{\detokenize{nonotes}}{
  \setbeameroption{hide notes}
}{
  \setbeameroption{show notes}
}
\setbeamertemplate{note page}[plain]
%\setbeameroption{hide notes}  % XXX

@@

\addtocounter{framenumber}{-1}

@@

\backupstart
{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/cambridge-university-botanic-garden}}
\begin{frame}[plain]
  \vspace{\textheight}\vspace{-1em}
  \centering
  \emph{Cambridge University Botanic Garden} \\
  \par
\end{frame}}

@@

\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}
\backupend

@@

  \begin{changemargin}{2.5em}{2.5em}
    We describe how the internal language of certain
    toposes, the associated small and big Zariski toposes of a scheme, can be
    used to give simpler definitions and more conceptual proofs of the basic
    notions and observations in algebraic geometry.
    % This is useful for studying schemes from a local or relative point of view.

@@

    A particularly interesting task is to internalize the
    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
    view, this construction should simply reduce to an intuitionistically sensible
    variant of the ordinary construction of the spectrum of a ring, but it turns
    out that this expectation is too naive and that a refined approach is
    necessary.

    % XXX: update
  \end{changemargin}
}

@@

  A sheaf~$\O_X$ of rings is \emph{local} if and only if all the
  stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
  sections~$\O_X(U)$ are local rings? This choice has a geometric meaning, but can
  also be motivated from a logical point of view: A sheaf of rings is local if
  and only if, from the point of view of the internal language of~$\Sh(X)$, it
  is a local ring.

@@

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\E$ \\
      \midrule
      object of~$\E$ & set/type \\
      morphism in~$\E$ & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      group object & group \\
      \bottomrule
    \end{tabular}
  \end{center}
}

\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
  \small
  Let~$X$ be a topological space. Then we recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  for open subsets~$U \subseteq X$ and formulas~$\varphi$. Write~``$\Sh(X)
  \models \varphi$'' to mean~$X \models \varphi$.
  \pause
  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
    U \models \varphi \wedge \psi &\Ll&
      \text{$U \models \varphi$ and $U \models \psi$} \\
    U \models \varphi \vee \psi &\Ll&
      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
    U \models \varphi \Rightarrow \psi &\Ll&
      \text{for all open~$V \subseteq U$: } 
    \text{$V \models \varphi$ implies $V \models \psi$} \\
    U \models \forall f \? \F\_ \varphi(f) &\Ll&
      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
      \varphi(f)$} \\
    U \models \exists f \? \F\_ \varphi(f) &\Ll&
      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
    $U_i \models \varphi(f_i)$}
  \end{array} \]
}

@@

    \item \begin{justify}For expositions of the internal language,
    see Chapters~D1 to~D4 of the Elephant, Chapter~VI of Moerdijk and Mac
    Lane's book, or Chapter~13 of
    \href{http://www.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{these
    lecture notes by Thomas Streicher}.\end{justify}
  \end{itemize}
}

\note{
  \begin{itemize}
    \item \begin{justify}The rules are called \emph{Kripke--Joyal semantics} and can be
    formulated over any topos (not just sheaf topoi). They are not all
    arbitrary: Rather, they are very finely concerted to make the crucial
    properties about the internal language (see next slide) true.\end{justify}

@@

    \item Negation is defined as
    $\neg\varphi :\equiv (\varphi \Rightarrow \bot)$.

    \item \begin{justify}The alternate definition ``$U \models \varphi \vee \psi :\Leftrightarrow
    \text{$U \models \varphi$ or $U \models \psi$}$'' would not be local (cf.
    next slide).\end{justify}
  \end{itemize}
}

@@

\note{\fontsize{8pt}{9.6}\selectfont
  \begin{itemize}
    \item One can simplify the rules for often-occuring special cases:
    \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
      U \models \forall s\?\F\_ \forall t\?\G\_ \varphi(s,t)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$,} \\
      && \text{sections~$s \in \F(V)$, $t \in \G(V)$:} \\
      &&\qquad\qquad V \models \varphi(s,t) \\\\
      U \models \forall s\?\F\_ \varphi(s) \Rightarrow \psi(s)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$, sections~$s \in \F(V)$:} \\
      &&\qquad\qquad \text{$V \models \varphi(s)$ implies $V \models \psi(s)$} \\\\
      U \models \exists!s\?\F\_ \varphi(s)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$,} \\
      &&
      \text{there is exactly one section~$s \in \F(V)$ with:} \\
      &&\qquad\qquad V \models \varphi(s)
    \end{array} \]
  \end{itemize}
}

\note{
  \begin{itemize}
    \item \begin{justify}One can extend the language to allow for \emph{unbounded}
    quantification ($\forall A$ vs.~$\forall a \in A$), by Shulman's
    \emph{stack semantics}. This is needed to formulate
    universal properties internal to~$\Sh(X)$, for instance.\end{justify}

    \item \begin{justify}One can further extend the language to be able to talk
    about locally internal categories over~$\Sh(X)$ (in the sense of
    Penon, see for instance the appendix
    of Johnstone's first topos theory book): Then one can do category theory internal
    to~$\Sh(X)$ using the internal language.

    This specific approach is, as far as I am aware, original work. But of
    course, internal category theory has been done for a long time, see for
    instance the Elephant and also Chapman and Rowbottom's \emph{Relative
    Category Theory and Geometric Morphisms: A Logical Approach}.\end{justify}
  \end{itemize}
}

@@

\note{\scriptsize%
  \vspace{-0.5em}%
  \begin{center}\large\bf Why is constructive mathematics
  interesting?\end{center}
  \vspace{-1em}
  \begin{itemize}
    \item The internal logic of most toposes is constructive.

@@

\note{\scriptsize%
  \vspace{-0.5em}%
  \begin{center}\large\bf Three rumours about constructive mathematics\end{center}
  \vspace{-1em}

  \begin{enumerate}
    \item \begin{justify}There is a false rumour about constructive mathematics, namely that
    the term \emph{contradiction} is generally forbidden. This is not the
    case, one has to distinguish between
    \begin{itemize}
      \item\scriptsize a true proof by contradiction: ``Assume~$\varphi$ were false.
      Then \ldots, contradiction. So~$\varphi$ is in fact true.''
    \end{itemize}
    which constructively is only a proof of the weaker
    statement~$\neg\neg\varphi$, and
    \begin{itemize}
      \item\scriptsize a proof of a negated formula: ``Assume~$\psi$ were true. Then
      \ldots, contradiction. So~$\neg\psi$ holds.''
    \end{itemize}
    which is a perfectly fine proof of~$\neg\psi$ in constructive mathematics.\end{justify}

@@

\note{
  \begin{enumerate}
    \item[3.]\scriptsize \begin{justify}There is one last false rumour about constructive mathematics: Namely
    that most of mathematics breaks down in a constructive setting. This is
    only true if interpreted naively: Often, already very small changes to the
    definitions and statements (which are classically simply equivalent
    reformulations) suffice to make them constructively acceptable.

    In other cases, adding an additional hypothesis, which is classically
    always satisfied, is necessary (and interesting). Here is an example: In
    constructive mathematics, one can not show that any inhabited subset of the
    natural numbers possesses a minimal element. [One can also not show the
    negation -- recall the previous false rumour.] But one can show (quite
    easily, by induction) that any inhabited and \emph{detachable} subset
    of the natural numbers possesses a minimal element. A subset~$U \subseteq
    \NN$ is detachable iff for any number~$n \in \NN$, it holds that~$n
    \in U$ or~$n \not\in U$.

@@

  \begin{justify}\scriptsize References about constructive mathematics include:
  \begin{itemize}
  \item Bridges. Constructive Mathematics.
  \item van Dalen. Intuitionistic logic.
  \item Troelstra and van Dalen. Constructivism in Mathematics: An Introduction.
  \end{itemize}
  \href{http://math.andrej.com/category/constructive-math/}{Andrej Bauer's
  blog} is also very informative.\end{justify}
}

@@

@@

\begin{frame}\frametitle{Building a dictionary}
  \sloganwithoutborder{\hil{Understand notions of algebraic geometry
  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \scalebox{0.93}{\begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set/type \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      finite locally free sheaf & finite free module \\
      % coherent sheaf & coherent module \\
      tensor product of sheaves & tensor product of modules \\
      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
      sheaf of Kähler differentials & module of Kähler differentials \\
      sheaf of rational functions & total quotient ring of~$\O_X$ \\
      dimension of $X$ & Krull dimension of~$\O_X$ \\
      \bottomrule
    \end{tabular}}
  \end{center}

@@

\note{
  \begin{block}{Proposition}
    If every inhabited subset of the natural numbers has a minimum, then the
    law of excluded middle holds. (So in constructive mathematics, one cannot
    prove the natural numbers to be complete in this sense.)
  \end{block}

  \begin{block}{Proof}
    Let~$\varphi$ be an arbitrary formula. Define the subset
    \[ U \defeq \{ n \in \NN \,|\, n = 1 \vee \varphi \} \subseteq \NN, \]
    which surely is inhabited by~$1 \in U$. So by assumption, there exists a
    number~$z \in \NN$ which is the minimum of~$U$. We have
    \[ z = 0 \quad\vee\quad z > 0 \]
    (this is constructively not trivial, but can be proven by induction).

    If~$z = 0$, we have~$0 \in U$, so~$0 = 1 \vee \varphi$, so~$\varphi$
    holds.

    If~$z > 0$, then~$\neg\varphi$ holds: Because if~$\varphi$ were
    true, zero would be an element of~$U$, contradicting the minimality of~$z$.
  \end{block}
}

\note{
  \begin{block}{Proposition}
    The partially ordered set
    \[ \widehat\NN \defeq \{ A \subseteq \NN \,|\, \text{$A$ inhabited and upward
    closed}
    \} \]
    is the least partially ordered set containing~$\NN$ and possessing minima
    of arbitrary inhabited subsets.
  \end{block}

  The embedding $\NN \hookrightarrow \widehat\NN$ is given by
  \[ n \in \NN \longmapsto {\uparrow}(n) \defeq \{ m \in \NN \,|\, m \geq n \}. \]

  \begin{block}{Proof}
    If~$M \subseteq \widehat\NN$ is an inhabited subset, its minimum is
    \[ \min M = \bigcup M \in \widehat\NN. \]
    The proof of the universal property is straightforward.
  \end{block}
}

\note{
  \begin{block}{External translation (see Mulvey's \emph{Intuitionistic algebra
  and representations of rings})}
    Let~$X$ be a topological space and consider the constant sheaf~$N$ with
    $\Gamma(U, N) = \{ f : U \to \NN \,|\ \text{$f$ continuous} \}$.
    Internally, the sheaf~$N$ plays the role of the ordinary natural numbers.
    Then there is an one-to-one correspondence:
    \begin{enumerate}
      \item Let~$A \hookrightarrow N$ be a subobject which is inhabited
      and upward closed from the internal point of view. Then
      \[
        x \longmapsto \inf\{ n \in \NN \,|\, n \in A_x \}
      \]
      is an upper semi-continuous function on~$X$.

      \item Let~$\alpha : X \to \NN$ be a upper semi-continuous function. Then
      \[ U \subseteq X \longmapsto \{ f : U \to \NN \,|\, \text{$f$
      continuous,\ \ $f \geq \alpha$ on~$U$} \} \]
      is a subobject of~$N$ which internally is inhabited and upward closed.
    \end{enumerate}
  \end{block}

  Under the correspondence, locally \emph{constant}
  functions map exactly to the [image of the] \emph{ordinary} internal natural numbers
  [in the completed natural numbers].
}

\note{
  \begin{itemize}
    \item Here is an explicit example of a completed natural number which is
    not an ordinary natural number: Let~$X =
    \Spec k[X]$ and~$\F = k[X]/(X-a)^\sim$. The rank of~$\F$ is~$1$ at~$a$ and
    zero elsewhere. It corresponds to the internal completed natural number
    \begin{multline*}
      z \defeq \min \{ n \in \NN \,|\, \speak{$\F$ can be generated by~$n$ elements} \} = \\
        \min\{ n \in \NN \,|\, n \geq 1 \vee \speak{the element $(X-a)$ of~$\O_X$
        is invertible} \}.
    \end{multline*}
    We have the internal implications
    \begin{align*}
      \Sh(X) \models \phantom{\neg}\,\speak{$(X-a)$ inv.} \Rightarrow z = 0\phantom{,} \\
      \Sh(X) \models \neg\,\speak{$(X-a)$ inv.} \Rightarrow z = 1,
    \end{align*}
    but we do \emph{not} have
    \begin{align*}
      \Sh(X) \models \speak{$(X-a)$ inv.} \ \vee\ \neg\,\speak{$(X-a)$ inv.},
    \end{align*}
    which would imply
    \begin{align*}
      \Sh(X) \models z = 0 \vee z = 1,
    \end{align*}
    i.\,e. the false statement that~$\F$ is locally free (of ranks~$0$ resp.~$1$).
  \end{itemize}
}

\note{
  \begin{itemize}
    \item \begin{justify}Here is a constructive proof of the statement that
    finitely generated vector spaces, for which the minimal number of
    generators is an actual natural numbers, are free:

    By assumption, the minimal number~$n \in \NN$ of generators for~$M$ exists.
    Let~$x_1,\ldots,x_n$ be a generating family of minimal length~$n$. We want
    to verify that it's linearly independent, so that it constitutes a basis.

    Let~$\sum_i \lambda_i x_i = 0$. If any~$\lambda_i$
    were invertible, the shortened family $x_1,\ldots,x_{i-1},$ $x_{i+1},\ldots,x_n$
    would also generate~$M$. By minimality of~$n$, this is not possible. So
    each~$\lambda_i$ is not invertible. By the field assumption on~$A$, it
    follows that each~$\lambda_i$ is zero.\end{justify}

    \item \begin{justify}In constructive mathematics, one can not show that
    every finitely generated vector space over a field admits a finite basis.
    (Exercise: Prove this by showing that this would imply the law of excluded
    middle.)
    This is not because the space might strangely turn out to be
    infinite-dimensional, but merely because one may not be able to explicitly
    exhibit a finite basis.\end{justify}
  \end{itemize}
}

@@

  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
  \begin{itemize}
  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}.
  \end{itemize}
  Note: $M$ and $\underline{M}$ share all first-order properties.
  \pause
  \vfill

@@

\note{\justifying
  The concept of a \emph{prime filter} is a direct axiomatization of what you
  expect the complement of a prime ideal to fulfil. In classical logic,
  complementation gives a bijection between the prime filters and the prime
  ideals of a ring.

  Prime filters are important in constructive mathematics because localizing
  them gives rise to local rings. In contrast, localizing a ring at the
  complement of a prime ideal doesn't usually result in a local ring.

  To construct the universal localization of~$A$, one doesn't pick a particular
  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
  in~$\Sh(\Spec A)$.
  \par
}

@@

\begin{frame}\frametitle{The curious role of affine open subsets}
  \hil{Question:} Why do the following identities hold, for quasicoherent
  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
  \begin{align*}
    \E/\F(U) &= \E(U) / \F(U) \\
    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
  \end{align*}
  \pause
  \hil{A calculation:}
  \begin{align*}
    M^\sim \otimes_{\O_U} N^\sim &=
    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
    &= (\ull{M \otimes_A N})[\F^{-1}] =
    (M \otimes_A N)^\sim.
  \end{align*}
  \pause
  \hil{Answer:} Because localization commutes with quotients, tensor products,
  torsion submodules (sometimes), \ldots
\end{frame}

@@

  Then~$\E$ is quasicoherent
  if and only if, internally to~$\Sh(X)$,
  \begin{quote}\textnormal{$\E[f^{-1}]$ is a $\Box_f$-sheaf for any~$f : \O_X$, \\[0.3em]
  \qquad\qquad where~$\Box_f\varphi \defeqv (\text{$f$ invertible} \Rightarrow \varphi)$.}
  \end{quote}
  \pause

@@

\note{\justifying
  The sheaf condition and the sheafification functor can be described purely
  internally. An object~$M$ is \emph{separated} with respect to~$\Box$ if
  and only if, from the internal point of view,
  \[ \forall x,y : M\_\ \Box(x = y) \Rightarrow x = y. \]
  It is a \emph{sheaf} with respect to~$\Box$, if furthermore
  \[ \forall K \subseteq M\_\
    \Box(\exists x : M\_\ K = \{ x \}) \Longrightarrow
    \exists x : M\_\ \Box(x \in K). \]

@@

\newcommand{\idiamond}{{\usebeamercolor[fg]{item}{\boldsymbol{\Box}}}}
\newcommand{\gdiamond}[1]{\textcolor{gray}{\boldsymbol{\Box}(}#1\textcolor{gray}{)}}

\begin{frame}\frametitle{The $\Box$-translation}
  Let~$\E_\Box \hookrightarrow \E$ be a subtopos given by a local
  operator. Then
  \[ \E_\Box \models \varphi \qquad\text{iff}\qquad
    \E \models \varphi^\Box\only<2->{.}\only<1>{,}\]
  \only<1>{where the translation~$\varphi \mapsto \varphi^\Box$ is given by:
  \begin{align*}
    (s = t)^\Box &\defeqv \idiamond(s=t) \\
    (\varphi \wedge \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \wedge \psi^\Box} \\
    (\varphi \vee \psi)^\Box &\defeqv \idiamond(\varphi^\Box \vee \psi^\Box) \\
    (\varphi \Rightarrow \psi)^\Box &\defeqv \gdiamond{\varphi^\Box \Rightarrow \psi^\Box} \\
    (\forall x\?X\_ \varphi(x))^\Box &\defeqv \gdiamond{\forall x\?X\_ \varphi^\Box(x)} \\
    (\exists x\?X\_ \varphi(x))^\Box &\defeqv \idiamond(\exists x\?X\_ \varphi^\Box(x))
  \end{align*}}%
  \only<2->{%
    Let~$X$ be a scheme. Depending on~$\Box$,~$\Sh(X)
    \models \Box\varphi$ means that~$\varphi$ holds on \ldots
    \begin{itemize}
      \item \ldots{} a dense open subset.
      \item \ldots{} a schematically dense open subset.
      \item \ldots{} a given open subset~$U$.
      \item \ldots{} an open subset containing a given closed subset~$A$.
      \item \ldots{} an open neighbourhood of a given point~$x \in X$.
    \end{itemize}

    \pause
    Can tackle the question~``$\varphi^\Box \stackrel{?}{\Rightarrow} \Box\varphi$'' logically.
  }
\end{frame}

\note{\justifying
  The~$\Box$-translation is a generalization of the \emph{double negation
  translation}, which is well-known in logic. The double negation translation
  has the following curious property: A formula~$\varphi$ admits a classical
  proof if and only if the translated formula~$\varphi^{\neg\neg}$ admits an
  intuitionistic proof.
